Note on Boundary condition and choice of grid

### 2024-07-08 write2/manuscript_part2_2024_hw.tex
comments and edits on manuscript from Jun 25th 2024 by Cedrick

\author{Hauke Wurps and Cedrick Ansorge}

@@

\begin{abstract}
  % Introduction % Theoretical profile
  Turbulent Ekman flow is a canonical flow representation of the atmospheric boundary layer, characterized by a logarithmic increase of the wind speed near the surface and a turning of the wind vector aloft. In part I of this work, we derived a formulation of the mean wind velocity profiles of the stationary turbulent Ekman flow based on direct numerical simulation (DNS). Here, we explore the extrapolation of these profiles to atmospheric Reynolds numbers using large-eddy simulation (LES).
  % Case desription and setup
  We compare theoretical profiles of the wind vector for low Reynolds-number turbulent Ekman flow to results from DNS and LES. This necessitates the inclusion of viscous effects, usually neglected in LES, along with modifications to the standard bottom boundary condition.
  % Results and Discussion
  Analysis of the grid-convergence for higher resolution unveils a convergence of LES data towards the theoretical profiles for both intermediate and high Reynolds numbers. Our LES thus confirm that--within the logarithmic layer (the lowest 10\,\% of the boundary layer)--- the turning of the wind vector rotates by approximately one-third of its total rotation across the entirety of the boundary layer.
  % Conclusions
  Such agreement of our theoretical formulation and LES data raises our confidence in the underlying assumptions of the theory and thus reinforces the utility of the theoretical profiles as a reference for intermediate and a quasi-reference for higher Reynolds number simulations.
%

@@

% \todo{start introduction rather with TEF than LES (rather subject than method)}
The interaction of atmospheric boundary layer  (ABL) flow with its lower boundary condition, the surface, is a defining property of the ABL \citep{stull:1988}. Relevant aspects of the interaction with the boundary not only include its heterogeneity \citep{avissar:MWR1989,giorgi:RG1997,claussen:BM1991,garratt:BM1990} and roughness \citep{monin:ARF1970,brutsaert:WRR1975,raupach:AMR1991,kostelecky:JFM2024}, but also a wind veer in favor of the pressure gradient force when the surface is approaced \citep{ekman:AMA1905}. An important tool with capabilities to model the flow-surface interaction is Large-Eddy Simulation (LES). Indeed, LES is widely used to model turbulent flow in the atmospheric boundary layer (ABL) \citep{stoll2020large}. This includes various applications, such as simulating wind turbine and wind farm wakes, and interactions within wind farm clusters \citep{porte2011large,mehta2014large,breton2017survey}. Furthermore, LES is applied in complex environments like mountainous or urban areas \citep{stoll2020large,garcia2018predictive} and to investigate the dispersion of pollution \cite{han2018large}. These applications typically focus on the lower segment of the ABL on the rotating earth, known as the Prandtl or surface layer. In this layer, the vertical wind speed profile commonly exhibits a logarithmic increase, followed by a prominent change of wind direction within the Ekman layer above. Accurate characterization of how wind speed and direction vary with height is of great importance for wind-power forecasting and projection \citep{optis2014moving}.

@@

\begin{align}\label{red}
 Re_D = \sqrt{2}\frac{G}{\sqrt{\nu f}}
\end{align}

@@

Part I of this publication, presents a framework to predict wind direction and speed across the turbulent Ekman flow, drawing from the works of \cite{csanady1967resistance}, \cite{tennekes1973logarithmic}, and \cite{spalart1989theoretical}. As \cite{spalart1989theoretical}, we use $G$, $f$, and $\nu$ to derive the friction velocity $u_\star$ and the angle $\alpha_\star$ between geostrophic wind and shear stress at the wall. While earlier descriptions of the turbulent Ekman layer were limited to specific parts of the boundary layer, we derive mean velocity profiles covering the entire ABL with due regard of changes in the Reynolds number. We have shown that the theoretical velocity profiles are in excellent agreement with DNS for intermediate Reynolds numbers.

However, it remains uncertain, if the extrapolation of the above framework towards atmospheric $\RE$ is valid. The substantial separation of scales in the ABL renders direct numerical simulation (DNS) of the entire turbulent flow at scale impossible. Within the framework of LES, only the larger eddies are resolved, that contain the main share of kinetic energy, while turbulent mixing below the subfilter-scale is parameterized. This technique allows for the simulation of the ABL at very high (atmospheric) $\RE$.

The accurate representation of mean profiles of velocity, i.e. of first-order statistics of the wind field, is commonly assumed when using LES in a predictive manner  \citep{fedorovich2004convective}. Here, we compare LES outcomes to theoretical solutions to determine the validity of this assumption under different conditions. As LES introduces the grid size as an additional parameter (often understood in relation to a filter scale, cf. \cite{pope2004ten}), we conduct a thorough analysis of its impact on simulation results. \cite{esau2004simulation} and \cite{jiang2018large} have executed similar LES analyses of neutral Ekman layers, however, they introduce an additional parameter by selecting an arbitrary roughness length $z_0$, whereas we derive the parameter $z_0$ corresponding to a fluid with a kinematic viscosity of $\nu = 1.5\cdot10^{-5}$ over flat surface.

Here, we examine three different Reynolds numbers $Re_D$: $1600$, $1.5\cdot10^5$, and $10^6$. While the case $Re_D=1600$ is also investigated by DNS (cf. part 1 of this study), $Re_D=1.5 \cdot 10^5$ and $10^6$ correspond to the scale separation found in typical atmospheric boundary layers. For $Re_D=1600$, we compare DNS and LES results using the exact same forcing and settings. This is the first comparison of this kind for the turbulent Ekman flow. The LES results converge to the DNS results with decreasing grid cell size. This approach raises our confidence in the LES results to a new level for the direct comparison with a resolving simulation. Furthermore, it allows us to extend the insight from DNS to atmospheric scale where an approach through DNS is clearly infeasible for computational constraints. We observe a clear $\RE$ dependency in the LES results, matching the predictions of the theoretical framework.

@@

In this chapter, we introduce the velocity profile for the turbulent Ekman layer developed in part I. 
%Definitions
We align the coordinate Ox with the surface stress, Oz points normal to the surface and normal to Ox, and Oy in the span-wise direction, normal to Oxz.
%The mean velocity profiles for all $\RE$ exhibit a dual scaling nature: in proximity to the lower boundary, the velocity profiles collapse when scaled using
In proximity to the lower boundary, the boundary layer scales in viscous (inner) units, denoted by the index ``$+$'' ($z^+ = zu_\star/\nu$, $U^+=U/u_\star$, where $U$ is the velocity component in x-direction and $u_\star = \sqrt{\nu \p_z U}$ the friction velocity). From the top perspective, the profiles converge when scaled using outer units, denoted by the index ``$-$'' ($z^-=z/\delta$, $U^-=U/G$, where $\delta=u_\star/f$ is the boundary-layer depth scale, $f$ the Coriolis parameter and $G$ the geostrophic wind speed). 
At the lower boundary, the x-axis of the inner units aligns with the shear stress, while the x-axis of the outer units aligns with the geostrophic wind. The angle between both axes is denoted as $\alpha_\star$, the surface veer of the wind across the boundary layer. 
%
Profiles of the velocity components are studied in three different layers reflecting the change of dominant balance when moving away from the surface. These are
(i) the Ekman layer denoted $(\cdot)_\text{Ek}$ (Sec.~\ref{sec:ekmanlayer}), 
(ii) the viscous sub-layer, denoted by an index $(\cdot)_\text{visc}$, 
and (iii) the logarithmic layer denoted by $(\cdot)_\text{log}$
We match $U_\text{visc}$ and $U_\text{log}$ according to their respective formulations and combine them to $U_\text{inner}$ (Sec.~\ref{sec:innerlayer}).
Inner and outer profiles are then matched by a weighting transfer function based on the error function. 
%The span-wise component of the velocity is separated into two layers, namely the inner layer $V_\text{inner}$, and the Ekman layer $V_{EK}$.

% \subsection{Drag-Law}

The geostrophic drag, i.e. normalized friction $Z\equiv u_\star/G$, and $\alpha_\star$ form the basis of the boundary-layer scaling, and they are estimated using the semi-empirical drag-law introduced by \cite{spalart1989theoretical}, which describes them as functions of only the Reynolds number:
\begin{subequations}\label{eqn:drag}
	\begin{align}
		\frac{G}{u_\star}\cos\phi^\star &= \frac{1}{\kappa}\log Re_\tau + C - A_r, \\
		\sin\phi^\star &= A_i\frac{u_\star}{G},\\
		\alpha_\star &= \phi^\star - \frac{C_5}{Re_\tau},\\
		Re_\tau &= \frac{u_\star^2}{\nu f} \left( = \frac{1}{2} Re_{D}^2 \frac{u_{\star}^2}{G^2}\right),
	\end{align}
\end{subequations}
where we use $\kappa = 0.416$, $A_r = 4.80$, $A_i = -5.57$, $C = 5.4605$, $C_5 = -57.8$ based on the DNS data available (cf. Part I, \cite{ansorge2014global}).

% \subsection{Total profile}

\subsection{Ekman layer}
\label{sec:ekmanlayer}
The outer layer of the ABL is characterized by a triadic balance between turbulent flux, pressure gradient and the Coriolis; the vertical change of the Coriolis force causes a pronounced height dependence of the wind direction resulting in the Ekman spiral \citep{ekman:AMA1905}. The classic solution employs Ekman dynamics down to the surface---and thus also in the surface layer, where this is a rather poor representation of turbulent mixing. Here, we model the surface layer by a lower boundary condition for the Ekman spiral determined from DNS:
\begin{subequations}
  \begin{align}
    U_{EK} &= G + Ae^{-\tilde{z}}\cos\tilde{z},\\
		V_{EK} &= - Ae^{-\tilde{z}}\sin\tilde{z},
  \end{align}
\end{subequations}
where the x-axis is aligned with the geostrophic wind and $A = 8.4u_\star$, $\tilde{z} = (z-z_r)/D_E$, $z_r = 0.12\delta$, and $D_E = 3\delta/4\pi\approx 0.24\delta$. The transition from the logarithmic layer to the Ekman layer is located at $z^{-}=0.28- 2.25 Re_{D}^{-1/2}$ with a transition scale of $\sigma_T=2$ for the stream-wise velocity.

\subsection{Inner and viscous layers}
\label{sec:innerlayer}

In the viscous sublayer, the shear-aligned velocity $U^{\alpha_\star}$ is described by the law of the wall $U^{\alpha_\star+} = z^+$ and the span-wise velocity $V^{\alpha_\star}$ is zero by choice of the reference frame 
(the index $\alpha_\star$ indicates the alignment with the direction of the shear stress). Around $z^+=5$, the velocity begins to deviate from linearity, and the buffer layer forms the transition from the viscous to the  logarithmic layer. From the surface up to the buffer layer, the stream-wise velocity is described by
\begin{subequations}
  \begin{align}
    U_\text{buffer}^{\alpha_\star+} = \frac{z^+}{1+c_1(z^+)^2} + (c_2z^+-a_\text{match})\frac{1+\tanh[0.2(z^+-22)]}{2}+c_3e^{-c_4(z^+-22)^2}.
    \label{eqn:ubuffer}
  \end{align}
With $c_1 = 0.00185$, $c_2 = 0.195$, $c_3 = 0.4$, $c_4 = 0.35$. The coefficient $a_{match}=3.5727$ is chosen to match the u-profile in the logarithmic layer above at $z^+=40$, such that 
\begin{align}
  U_\text{inner}^{\alpha_\star+} = \begin{cases}
	  U_\text{buffer}^{\alpha_\star+}, & z^+ \le 40.\\
		\kappa^{-1}\log z^+ + C &   z^+>40.
	\end{cases}
  \label{eqn:log-law}
\end{align}
\end{subequations}
with the von-Kármán constant $\kappa=0.416$, and $C=5.4605$ as in the estimation of the geostrophic drag in Eq.~(\ref{eqn:drag}).

For the span-wise velocity, the boundary conditions and Ekman dynamics imply that 
\begin{subequations}
\begin{align}
V^{\alpha_\star}_\text{visc} \delta^+\sim G f_V(z^+),
\end{align} where $f_V$ is a universal, non-dimensional function (cf. Part I). Above the viscous 
layer, scaling arguments are scarce, but there is evidence for a logarithmic scaling of also the 
span-wise velocity component. We  hence use  
\begin{align}
  V^{\alpha_\star}_\text{inner} = \left\{ \begin{matrix}
      V^{\alpha_\star}_\text{visc} &=& \dfrac{G}{\delta^+}v_{ref}(w_vz^+-1+exp[-w_vz^+])  & z^+ \le 10\\ 
      V_\text{log}^{\alpha_\star}        &=& \dfrac{G}{\delta^+}(a_{log} + b_{log}\log(z^+) + c_{log}z^+) & z^+>10. 
  \end{matrix}\right.
\end{align}
\end{subequations}
with $v_\text{ref} = 18.85$ and $w_v = 0.2353$ leads to excellent agreement with the DNS data. The coefficients $a_\text{log}, b_\text{log}$, and $c_\text{log}$ are determined by (i)-(ii) a smooth transition to $V_\text{visc}$ at $z^+=10$, and (iii) the condition $V_\text{log}(z^-=0.3) = V_\text{EK}(z^-=0.3) =: v_{03}$ (cf. Part I). For the Reynolds-number dependency of the Ekman-layer profiles, $V_\text{Ek}(z^-=0.3)$ depends on $Re$, such that the coefficients exhibit a weak dependence on $Re$.

\subsection{Transition from inner to outer profiles, synopsis}

A smooth transition between consecutive layers is achieved using a transfer function:
\begin{align}\label{eqn:error}
  w_{*} = \frac{1}{2}\left(\textrm{erf}\left[\sigma_T\log\left(\frac{z}{z_{T}}\right)\right]+1\right),
\end{align}
where $erf$ is the error function, $\sigma_T$ is a transition scale that defines the width of the transition and $z_{T}$ is the height of the transition, where the upper and the lower layer equally contribute to the velocity ($w_{*}(z_{T})=0.5$). The inner profiles $U_\text{inner}$ and $V_\text{inner}$ are blended to the Ekman profile using eq.~(\ref{eqn:blend}) with $\sigma_T = 2$ and $z_T^- = 0.28- 2.25\sqrt{1/Re_D}$.

\begin{subequations}\label{eqn:blend}
  \begin{align}
    U &= (1-w_{outer})U_{inner} + w_{outer}U_{EK},\\
	  V &= (1-w_{outer})V_{inner} + w_{outer}V_{EK}.
	\end{align}
\end{subequations}

@@

The velocity profiles are given for two Reynolds numbers in Fig. \ref{single_ekman}. 
In the viscous sublayer, the shear-aligned component $U^+$ increases linearly with height up 
$z^+\approx5$. In between $5 \lesssim z^+ \lesssim 30$, we find the buffer layer with a transition from the linear law-of-the-wall to the logarithmic law. Above $z^+\approx30$, the logarithmic layer begins. Then, $U^+$ increases logarithmically up to the Ekman layer and reaches its supergeostrophic maximum. Above, it decreases to its free-stream geostrophic value. The depth of the logarithmic layer increases with $\RE$. The spanwise component $V^+$ remains close to zero up to the middle of the logarithmic layer, where the transition to the Ekman layer takes place ($z^-\approx0.28-2.25\sqrt{1/Re_D}$). The profiles of $V^+$ of all Reynolds numbers have a similar shape but are shifted in $z^+$. The similar $V^+$ and the growth of $U^+$ leads to a smaller angle $\alpha_\star$ between surface shear stress and geostrophic wind for higher $\RE$, which is visible in the hodograph.

@@

\begin{table*}
	\centering
	\caption{Parameters of the simulated cases. $f$, $\nu$, and $G$ are input parameters and define the Reynolds numbers $Re_D$ and $Re_\tau$, while $u_\star$, $\alpha_\star$, and $\delta = u_\star/f$ are resulting properties of the flow}
	\begin{tabular}{lcccccccc}
          \toprule  
	  Name & $Re_D$ & $Re_\tau$ & $f$ [s$^{-1}$] & $\nu$ [m$^2$s$^{-1}$] & $G$ [ms$^{-1}$] & $u_\star$ [ms$^{-1}$] & $\alpha_\star$ [$^\circ$]& $\delta$ [m] \\
          \midrule
	  Re1 & $1.6\times10^3$ & $3.0\times10^3$ &  &  & 0.0438 & 0.00211 & 16.8 & 21.1 \\
	  Re2 & $1.5\times10^5$ & $7.3\times10^6$ & $10^{-4}$ & $1.5\times10^{-5}$ & 4.108 & 0.1048 & 8.5 & 1048\\
	  Re3 & $1\times10^6$ & $2.2\times10^8$ &  &  & 27.39 & 0.5785 & 7.0 & 5785 \\
          \bottomrule
	\end{tabular}
	\label{simulation_parameters}
\end{table*}
An incompressible, turbulent Ekman flow over a hydrodynamically smooth surface is simulated using the Parallel Large Eddy Simulation model (PALM, \citealp{maronga2020overview}). We study three different Reynolds numbers Re$_D = 10^3;1.5\times 10^5;10^6$, hereafter Re1, Re2, and Re3, respectively (Tab.~\ref{simulation_parameters}).
%
The domain is rotating around the z-axis with an angular velocity corresponding to the Coriolis parameter $f=10^{-4}s^{-1}$. The stratification of the flow is truly neutral, i.e., the potential temperature is constant across the whole domain. A constant boundary layer height on the order of $\delta=u_\star/f$ forms due to the balance between shear production and rotational suppression of turbulence. At the upper boundary, the flow is non-turbulent. Hence, a no-penetration boundary condition is used for the vertical velocity component while the horizontal components are prescribed according to geostrophic balance--a Dirichlet-type boundary condition. At the bottom, a constant-flux layer is assumed and Monin--Obukhov similarity theory (MOST) is used to calculate the surface momentum fluxes. The Navier-Stokes equations are integrated using a 3$^\text{rd}$-order low-storage Runge--Kutta method. For scalar advection a 5$^\text{th}$-order Wicker--Skamarock scheme is employed. The Poisson equation is solved using a direct fast Fourier transform (FFT). In LES, the turbulent transport on the subgrid scale (SGS) needs to be modeled by an SGS closure model. We use two kinds of SGS closures to assess their impact on the LES solution: a 1.5-order closure after \cite{deardorff1980stratocumulus} and a dynamic closure after \cite{heinz2008realizability}. For most of the simulations, the 1.5-order closure is used, since the dynamic closure needs more computational resources, but the simulations with $\Delta^-=200^{-1}$ are repeated using the dynamic closure.

@@

\subsection{Geostrophic drag and surface wind veer}
\label{bulk}

>>>>

\author{hauke-zrick}

@@

\begin{abstract}
  % Introduction % Theoretical profile
	Turbulent Ekman flow is the simplest representation of the atmospheric boundary layer that exhibits two key properties: the logarithmic increase of the wind speed near the surface and a turning of the wind vector aloft. In part I of this work, we derived a formulation of the mean wind velocity profiles of the stationary turbulent Ekman flow based on direct numerical simulation (DNS). Here, we explore the feasibility of extrapolating these profiles to atmospheric Reynolds numbers using large-eddy simulation (LES).
	% Case desription and setup
  We compare theoretical wind speed profiles of a low Reynolds-number turbulent Ekman flow to results from DNS and LES. This necessitates the inclusion of viscous effects, usually neglected in LES, along with modifications to the standard bottom boundary condition.
	% Results and Discussion
	Analysis of the grid resolution dependency showed a convergence of LES results towards the theoretical profiles for low and high Reynolds numbers. LES results confirmed that within the logarithmic layer (the lowest 10\,\% of the boundary layer), the turning of the wind vector contributes approximately one-third to the total rotation.
	% Conclusions
  The notable agreement of the theoretical formulation and LES  results raised our confidence in its reliability for high Reynolds numbers. This reinforces the utility of the theoretical profiles as a true reference for intermediate and a quasi-reference for higher Reynolds number simulations.
%

@@

\todo{start introduction rather with TEF than LES (rather subject than method)}
Large Eddy simulation (LES) is widely used to model turbulent flow in the atmospheric boundary layer (ABL) \citep{stoll2020large}. This includes various applications, such as simulating wind turbine and wind farm wakes, and interactions within wind farm clusters \citep{porte2011large,mehta2014large,breton2017survey}. Furthermore, LES is applied in complex environments like mountainous or urban areas \citep{stoll2020large,garcia2018predictive} and to investigate the dispersion of pollution \cite{han2018large}. These applications typically focus on the lower segment of the ABL on the rotating earth, known as the Prandtl or surface layer. In this layer, the vertical wind speed profile commonly exhibits a logarithmic increase, followed by a prominent change of wind direction within the Ekman layer above. Accurate characterization of how wind speed and direction vary with height is of great importance for wind-power forecasting and projection \citep{optis2014moving}.

@@

\begin{align}\label{red}
 Re_D = \frac{G}{\sqrt{\frac{1}{2}\nu f}}
\end{align}

@@

Part I of this publication, presents a theoretical framework to predict wind direction and speed across the turbulent Ekman flow, drawing from the works of \cite{csanady1967resistance}, \cite{tennekes1973logarithmic}, and \cite{spalart1989theoretical}. As \cite{spalart1989theoretical}, we use $G$, $f$, and $\nu$ to derive the friction velocity $u_\star$ and the angle $\alpha_\star$ between geostrophic wind and shear stress at the wall. While earlier descriptions of the turbulent Ekman layer were limited to specific parts of the boundary layer, we give a comprehensive theoretical description of the mean velocity profiles that covers the complete ABL and quantifies the Reynolds number. We have shown that the theoretical velocity profiles are in excellent agreement with DNS for low and intermediate Reynolds number.

However, it remains uncertain, if the extrapolation of the theory towards atmospheric $\RE$ is valid. The substantial separation of scales in the ABL renders direct numerical simulation (DNS) of the entire turbulent flow at scale impossible. Within the framework of LES, only the larger eddies are resolved, that contain the main share of kinetic energy, while turbulent mixing below the subfilter-scale is parameterized. This technique allows for the simulation of the ABL at very high (atmospheric) $\RE$.

When using LES in a predictive manner, a common assumption is the accurate reproduction of velocity profiles, i.e. first-order statistics of the wind field \citep{fedorovich2004convective}. Here, we compare LES outcomes to theoretical solutions to determine the validity of this assumption under different conditions. As LES inherently introduces the grid size as an additional (artificial) parameter, we conduct a thorough analysis of its impact on simulation results. \cite{esau2004simulation} and \cite{jiang2018large} have executed similar LES analyses of neutral Ekman layers, however, they introduce an additional parameter by selecting an arbitrary roughness length $z_0$, whereas we derive the parameter $z_0$ corresponding to a fluid with a kinematic viscosity of $\nu = 1.5\cdot10^{-5}$ over flat surface.

Here, we examine three different Reynolds numbers $Re_D$: $1600$, $1.5\cdot10^5$, and $10^6$. While the low-Re case ($Re_D=1600$) is also investigated by DNS (cf. part 1 of this study), $Re_D=1.5 \cdot 10^5$ and $10^6$ correspond to the scale separation found in typical atmospheric boundary layers. For the low-Re case, we compare DNS and LES results using the exact same forcing and settings. This is the first comparison of this kind for the turbulent Ekman flow. The LES results converge to the DNS results with decreasing grid cell size. This approach raises our confidence in the LES results to a new level for the direct comparison with a resolving simulation. Furthermore, it allows us to extend the insight from DNS to atmospheric scale where an approach through DNS is clearly infeasible for computational constraints. We observe a clear $\RE$ dependency in the LES results, matching the predictions of the theoretical framework.

@@

A universal velocity profile for the turbulent Ekman layer is developed in part I of this paper. In this chapter, we give a brief mathematical description.

%Definitions
The mean velocity profiles for all $\RE$ exhibit a dual scaling nature: in proximity to the lower boundary, the velocity profiles collapse when scaled using viscous (inner) units, denoted by the index ``$+$'' ($z^+ = zu_\star/\nu$, $U^+=U/u_\star$, where $U$ is the velocity component in x-direction). From the top perspective, the profiles converge when scaled using outer units, denoted by the index ``$-$'' ($z^-=z/\delta$, $U^-=U/G$, with $\delta=u_\star/f$ representing the boundary layer height). The outer and inner velocities $G$ and $u_\star$, respectively, are linked by a semi-empirical law \cite{spalart1989theoretical}. At the lower boundary, the x-axis of the inner units aligns with the shear stress, while the x-axis of the outer units aligns with the geostrophic wind. The angle between both axes is denoted by $\alpha_\star$.

\subsection{Drag-Law}

The geostrophic drag $Z \equiv u_\star/G$ and the angle between the shear stress and the geostrophic wind $\alpha_\star$ are two key parameters of the Ekman flow. They can be estimated using a semi-empirical drag-law based on \cite{spalart1989theoretical}, which describes them as functions of only the Reynolds number:
\begin{subequations}\label{drag}
	\begin{align}
		\frac{G}{u_\star}\cos\phi^\star &= \frac{1}{\kappa}\log Re_\tau + C - A_r, \\
		\sin\phi^\star &= A_i\frac{u_\star}{G},\\
		\alpha_\star &= \phi^\star - \frac{C_5}{Re_\tau},\\
		Re_\tau &= \frac{u_\star^2}{\nu f},
	\end{align}
\end{subequations}
with $Re_\tau = \frac{Re_D^2}{2}\frac{u_\star^2}{G^2}$ the turbulent Reynolds number, $\kappa = 0.416$, $A_r = 4.80$, $A_i = -5.57$, $C = 5.4605$, $C_5 = -57.8$. This law is in excellent agreement with DNS in the range $400\leq Re_D\leq 1\,600$ as demonstrated by \cite{ansorge2014global}.

\subsection{Total profile}

As frame of reference we use the coordinate system Oxyz with Ox in the direction of the surface stress, Oz normal to the surface, and Oy in the span-wise direction normal to Oxz. The profile of the stream-wise component of the velocity is seperated into three layers, which are the viscous layer $U_{visc}$, the logarithmic layer $U_{log}$, and the Ekman layer $U_{EK}$. $U_{visc}$ and $U_{log}$ are matched by their formulation and combined to $U_{inner}$. The span-wise component of the velocity is separated into two layers, namely the inner layer $V_{inner}$, and the Ekman layer $V_{EK}$. The smooth transition between consecutive layers is achieved using a transfer function:
\begin{align}\label{error}
  w_{*} = \frac{1}{2}\left(\textrm{erf}\left[\sigma_T\log\left(\frac{z}{z_{T}}\right)\right]+1\right),
\end{align}
where $erf$ is the error function, $\sigma_T$ is a transition scale that defines the width of the transition and $z_{T}$ is the height of the transition, where the upper and the lower layer equally contribute to the velocity ($w_{*}(z_{T})=0.5$).

\begin{subequations}\label{blend}
  \begin{align}
    U &= (1-w_{outer})U_{inner} + w_{outer}U_{EK},\\
	  V &= (1-w_{outer})V_{inner} + w_{outer}V_{EK}.
	\end{align}
\end{subequations}

\subsection{Profiles in Outer Layer}

In the outer layer, the vertical change of the Coriolis force causes a  pronounced height dependence of the wind direction resulting in the Ekman spiral \cite{ekman1905influence}. While the classic solution employs Ekman dynamics down to the surface --- and thus also in the surface layer, where the pure Ekman dynamics are a rather poor representation of turbulent mixing ---, we model the surface layer by a lower boundary condition for the Ekman spiral that we determine from DNS data:
\begin{subequations}
  \begin{align}
    U_{EK} &= G + Ae^{-\tilde{z}}\cos\tilde{z},\\
		V_{EK} &= - Ae^{-\tilde{z}}\sin\tilde{z},
  \end{align}
\end{subequations}
where the x-axis is aligned with the geostrophic wind and $A = 8.4u_\star$, $\tilde{z} = (z-z_r)/D_E$, $z_r = 0.12\delta$, and $D_E = 3\delta/4\pi\approx 0.24\delta$. The transition from the logarithmic layer to the Ekman layer is located at $z^-_T=0.3-120/Re_D$ \todo{in script at 0.28- 2.25*np.sqrt(1./re)} with a transition scale of $\sigma_T=2$ for the stream-wise velocity.

\subsection{Shear-Aligned Velocity}

In the viscous sublayer, the span-wise velocity is close to zero and the shear-aligned velocity is described by the law of the wall:

\begin{align}
  U^{\alpha_\star+} = z^+,
\end{align}
(the index $\alpha_\star$ indicates the alignment with the direction of the shear stress). Around $z^+=5$, the velocity is beginning to deviate from its linear profile and the buffer layer forms the transition between viscous layer and logarithmic layer. From the surface up to the buffer layer, the stream-wise velocity is described by
\begin{align}
  U_{visc}^{\alpha_\star+} = \frac{z^+}{1+c_1(z^+)^2} + (c_2z^+-a_{match})\frac{1+\tanh[0.2(z^+-22)]}{2}+c_3e^{-c_4(z^+-22)^2}.
\end{align}
With $c_1 = 0.00185$, $c_2 = 0.195$ \todo{adapt $c_2$?}, $c_3 = 0.4$, $c_4 = 0.35$. The coefficient $a_{match}=3.5727$ is chosen to match the u-profile in the logarithmic layer above at $z^+=40$.

The logarithmic region of the stream-wise velocity is 
\begin{align}\label{log-law}
  U_{\log}^{\alpha_\star+} = \frac{1}{\kappa}\log z^+ + C,
\end{align}
with the von-Kármán constant $\kappa=0.416$, and $C=5.4605$. The lower part of $U^{\alpha_\star+}$ is described by
\begin{align}
  U_{inner}^{\alpha_\star+} = \begin{cases}
	  U_{visc}^{\alpha_\star+}, & z^+<40.\\
		U_{log}^{\alpha_\star+}, & z^+>40.
	\end{cases}
\end{align}

The inner profile is then blended to the Ekman profile using eq. \ref{blend} with $\sigma_T = 2$ and $z_T^- = 0.28- 2.25\sqrt{1/Re_D}$.

\subsection{Span-wise Velocity}

In the viscous layer, the shear-aligned span-wise velocity is described as
\begin{align}
  V^{\alpha_\star}_{visc} = \frac{G}{\delta^+}v_{ref}(w_vz^+-1+exp[-w_vz^+]).
\end{align}
The choice of $v_{ref} = 18.85$ and $w_v = 0.2353$ leads to excellent agreement with the DNS data.

Above the viscous layer a log-like transition to the Ekman layer is modeled by
\begin{align}
  V_{log}^{\alpha_\star} = \frac{G}{\delta^+}(a_{log} + b_{log}\log(z^+) + c_{log}z^+).
\end{align}
A smooth transition to $V_{visc}$ at $z^+=10$ and the condition $V_{log}(z^-=0.3) = V_{EK}(z^-=0.3) =: v_{03}$ lead to
\begin{align}
  c_{log} &= \frac{v_{03}-v_{10} - d_{10}z_{10}\log(z_{03}/z_{10})}{z_{03}-z_{10}},\\
	a_{log} &= v_{10}-d_{10}z_{10}\log z_{10}+c_{log}z_{10}(1-\log z_{10}),\\
	b_{log} &= (d_{10}-c_{log})z_{10}.
\end{align}
Here, $d_{10}=4.01$, $v_{10}=27.3$. Because of the $\RE$ dependence of $V_{Ek}$, $v_{03}$ also depends on $\RE$. $V_{log}^{\alpha_\star}$ is blended into $V_{EK}$ using eq. \ref{blend} with $\sigma_T = 2$ and $z_T^- = 0.28- 2.25\sqrt{1/Re_D}$.

\subsection{Profile}

@@

The velocity profiles are given for two Reynolds numbers in Fig. \ref{single_ekman}. In the viscous sublayer, the shear-aligned component $U^+$ increases linearly with height up to the buffer layer between $z^+\approx5$ and $z^+\approx30$, where it transitions into the logarithmic layer. Then, $U^+$ increases logarithmically up to the Ekman layer and reaches its supergeostrophic maximum. Above, it decreases to its geostrophic value. The length of the logarithmic layer increases with $\RE$. The spanwise component $V^+$ remains close to zero up to the middle of the logarithmic layer, where the transition to the Ekman layer takes place ($z^-\approx0.28-2.25\sqrt{1/Re_D}$). The profiles of $V^+$ of all Reynolds numbers have a similar shape but are shifted in $z^+$. The similar $V^+$ and the growth of $U^+$ leads to a smaller angle $\alpha_\star$ between surface shear stress and geostrophic wind for higher $\RE$, which is visible in the hodograph.

@@

An incompressible, turbulent Ekman flow over a hydrodynamically smooth surface is simulated for three different Reynolds numbers $Re_D = 10^3;1.5\times 10^5;10^6$, hereafter $Re1$, $Re2$, and $Re3$, respectively. The input parameters are given in table \ref{simulation_parameters}.

\begin{table*}
	\centering
	\caption{Parameters of the simulated cases. $f$, $\nu$, and $G$ are input parameters and define the Reynolds numbers $Re_D$ and $Re_\tau$, while $u_\star$, $\alpha_\star$, and $\delta = u_\star/f$ are resulting properties of the flow}
	\begin{tabular}{lcccccccc}
          \toprule  
	  Name & $Re_D$ & $Re_\tau$ & $f$ [s$^{-1}$] & $\nu$ [m$^2$s$^{-1}$] & $G$ [ms$^{-1}$] & $u_\star$ [ms$^{-1}$] & $\alpha_\star$ [$^\circ$]& $\delta$ [m] \\
          \midrule
	  Re1 & $1.6\times10^3$ & $3.0\times10^3$ &  &  & 0.0438 & 0.00211 & 16.8 & 21.1 \\
	  Re2 & $1.5\times10^5$ & $7.3\times10^6$ & $10^{-4}$ & $1.5\times10^{-5}$ & 4.108 & 0.1048 & 8.5 & 1048\\
	  Re3 & $1\times10^6$ & $2.2\times10^8$ &  &  & 27.39 & 0.5785 & 7.0 & 5785 \\
          \bottomrule
	\end{tabular}
	\label{simulation_parameters}
\end{table*}

The domain is rotating around the z-axis with an angular velocity corresponding to the Coriolis parameter $f=10^{-4}s^{-1}$. The stratification of the flow is truly neutral, i.e., the potential temperature is constant for the whole domain. Nevertheless, the boundary layer height is not growing infinitely but results from a balance between shear production and rotational suppression of turbulence. At the upper boundary, sufficiently far aloft of the boundary-layer height, a no-penetration boundary condition is used and the horizontal components of the wind are geostrophic, which is a Dirichlet-type boundary condition. At the bottom, a constant-flux layer is assumed and Monin--Obukhov similarity theory (MOST) is used to calculate the surface momentum fluxes. The Navier-Stokes equations are integrated using a 3rd-order low-storage Runge-Kutta method. For scalar advection a 5th-order Wicker--Skamarock scheme is employed. The Poisson equation is solved using a direct fast Fourier transform (FFT). In LES, the turbulent transport on the subgrid scale (SGS) needs to be modeled by a dedicated model, the SGS model. We use two different SGS models to check whether the subgrid closure significantly influences the LES solution: a 1.5-order closure after \cite{deardorff1980stratocumulus} and a dynamic closure after \cite{heinz2008realizability}. For most of the simulations, the 1.5-order closure is used, since the dynamic closure needs more computational resources, but the simulations with $\Delta^-=200^{-1}$ are repeated using the dynamic closure.  A comprehensive description of the LES model is given by \cite{maronga2020overview}.

@@

\subsection{Bulk parameters}
\label{bulk}
